Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-12-20 paper/main.tex
adjust notation
clarify that this is a phenomenological approach

The physical interaction between the bacteria plays a crucial role for the overall dynamics of the
system.
The Mie potential~\cite{Mie1903} is capable of describing a wide range of interaction
types from softer ones that permit some level of overlap to rigid ones with steep ascends for
distance valuels corresponding to overlaps.
In contrast, a popular alternative is the Morse potential~\cite{Morse1929}, which is less
customizable due to the lower number of parameters and more restrictive mathematical formulation
but may therefore be advantagous in situations of sparse data.
These potentials are a good choice for a simple, interpretable interaction potential with
low number of parameters that can be estimated.

@@

% This metric can be used to compare the model output to datapoints obtained from microscopic images.
A common technique to investigate identifiability questions and provide insights into confidence
levels of the obtained results is the
\textit{profile-likelihood}~\cite{Murphy2000,Kreutz2013,Tnsing2023}.
It is able to provide confidence intervals which are invariant under parameter
transformations~\cite{Meeker1995}.
It builds on the so-called \textit{log-likelihood} which propagates uncertainties $\sigma_{ijkl}$ in
the datapoints.
Under the assumption of a constant, known uncertainty, i.e. $\sigma_{ijkl}=\sigma$, we can rewrite
the log-likelihood in terms of $L_\text{LS}$ via
\begin{align}
    L(\theta) &= \sum\limits_{ijkl}\left(\left(\frac{\nu_{ijkl}-\tilde{\nu}_{ijkl}}{\sigma_{ijkl}}\right)^2
    + 2 \log(\sigma_{ijkl})
    +\text{const}\right)\\
    &= \frac{L_\text{LS}(\theta)}{\sigma^2} + \text{const}
\end{align}
where we absorbed the terms $\log(\sigma)$ in the sum into the constant term.
With this simplification, it is important to pick a value for $\sigma$ such that the real
uncertainty is overestimated.
Furthermore, the preceding section (\nameref{subsec:position-extraction-algorithm}) has shown that
the error distribution of the underlying data follows a normal distribution which justifies the core
assumption of this framework.
We conclude that minimizing, the least squares value $L_\text{LS}$ under constant $\sigma$ is
equivalent to minimizing the \textit{log-likelihood} $L$ which is again equivalent to maximizing the
likelihood that the chosen parameters represent our model.

>>>>

The physical interaction between the bacteria plays a crucial role for the overall dynamics of the
system.
We follow a phenomenological approach thus aiming to fit our model to given datapoints and determine
the type of interaction in this way.
In order to accurately depict the physical size of the rod, our interactions will be repulsive for
short ranges $r<R_1+R_2$ and attractive at longer ranges.
The Mie potential~\cite{Mie1903} is capable of describing a wide range of interaction
types from softer ones that permit some level of overlap to rigid ones with steep ascends for
distance valuels corresponding to overlaps.
In contrast, a popular alternative is the Morse potential~\cite{Morse1929}, which is less
customizable due to the lower number of parameters and more restrictive mathematical formulation
but may therefore be advantagous in situations of sparse data.
These potentials are a good choice for a simple, interpretable interaction potential with
low number of parameters that can be estimated.

@@

% This metric can be used to compare the model output to datapoints obtained from microscopic images.
A common technique to investigate identifiability questions and provide insights into confidence
levels of the obtained results is the
\textit{profile-likelihood}~\cite{Murphy2000,Kreutz2013,Tnsing2023}.
It is able to provide confidence intervals which are invariant under parameter
transformations~\cite{Meeker1995}.
It builds on the so-called \textit{log-likelihood} which propagates uncertainties $\sigma_{ijkl}$ in
the datapoints.
Under the assumption of a constant, known uncertainty, i.e. $\sigma_{ijkl}=\sigma$, we can rewrite
the log-likelihood in terms of $L_\text{LS}$ via
\begin{align}
    L(\theta) &= \sum\limits_{ijkl}\left(\left(\frac{\nu_{ijkl}-\tilde{\nu}_{ijkl}}{\sigma_{ijkl}}\right)^2
    + 2 \log(\sigma_{ijkl})
    +\text{const}\right)\\
    &= \frac{L_\text{LS}(\theta)}{\sigma^2} + \text{const}
\end{align}
where we absorbed the terms $2\log(\sigma_{ijkl})$ in the sum into the constant term.
With this simplification, it is important to pick a value for $\sigma$ such that the real
uncertainty is overestimated.
Furthermore, the preceding section (\nameref{subsec:position-extraction-algorithm}) has shown that
the error distribution of the underlying data follows a normal distribution which justifies the core
assumption of this framework.
We conclude that minimizing, the least squares value $L_\text{LS}$ under constant $\sigma$ is
equivalent to minimizing the \textit{log-likelihood} $L$ which is again equivalent to maximizing the
likelihood that the chosen parameters represent our model.
